Effect of initial volume fraction on the collapse of granular columns in fluid

### 2018-08-26 paper/paper.tex
:scissors: Editing content on Potential Energy and Pressures

\subsection{Turbulence in Lattice Boltzmann Method}
Modelling fluids with low viscosity like water remains a challenge, necessitating very small values of \textit{h}, and/or $\tau$ very close to 0.5~\shortcite{He1997}. Turbulent flows are characterised by the occurrence of eddies with multiple scales in space, time and energy. In this study, the Large Eddy Simulation (LES) is adopted to solve for turbulent flow problems. The separation of scales is achieved by filtering of the Navier-Stokes equations, from which the resolved scales are directly obtained and unresolved scales are modelled by a one-parameter Smagorinski sub-grid methodology, which assumes that the Reynold's stress tensor is dependent only on the local strain rate~\shortcite{Smagorinsky1963}. The turbulent viscosity $\nu$ is related to the strain rate $S_{ij}$ and a filtered length scale `h' as follows:
\begin{align}
&\mathit{v}_{\mathit{t}}  =  (\mathit{S}_{c}\mathit{h})^{2}\overline{S}; \\
&\overline{S}  = \sqrt{\sum\limits_{\mathit{i,j}}{\tilde{S}_{\mathit{i,j}}\tilde{S}_{\mathit{i,j}}}}
\end{align}
where $\mathit{S}_{c}$ is the Smagorinski constant found to be close to 0.03~\shortcite{yu2005}.

The effect of the unresolved scale motion is taken into account by introducing an effective collision relaxation time scale $\tau_{t}$, so that the total relaxation time $\tau_{*}$ is written as:
\begin{align}
\tau_{*}=\tau + \tau_{t}
\end{align}
where $\tau$ and $\tau_{t}$ are respectively the standard relaxation times corresponding to the true fluid viscosity \textit{v} and the turbulence viscosity $\mathit{v}_{\mathit{t}}$, defined by a sub-grid turbulence model. The new viscosity $\mathit{v}_{*}$ corresponding to $\tau_{*}$ is defined as:
\begin{align}
& \mathit{v}_{*}=\mathit{v}+\mathit{v}_{\mathit{t}}=\frac{1}{3}(\tau+\tau_{t}-\frac{1}{2})\mathit{C}^{2} \Delta \mathit{t}  \\
& \mathit{v}_{\mathit{t}}=\frac{1}{3}\tau_{\mathit{t}}\mathit{C}^{2} \Delta \textit{t}
\end{align}
The Smagorinski model is easy to implement and the Lattice Boltzmann formulation remains unchanged, except for the use of a new turbulence-related viscosity $\tau_{*}$. The component $s_1$ of the collision matrix becomes $s_1 = \frac{1}{\tau+\tau_t}$.
%************************************************************************* %
\section{COUPLED LB - DEM FOR FLUID-PARTICLE INTERACTIONS}
The Lattice Boltzmann approach has the advantage of accommodating large particle sizes and the interaction between the fluid and the moving particles can be modelled through relatively simple fluid - particle interface treatments. Further, employing the Discrete Element Method (DE) to account for the particle/particle interaction naturally leads to a combined LB - DEM solution procedure. The Eulerian nature of the Lattice Boltzmann formulation, together with the common explicit time step scheme of both the Lattice Boltzmann and the Discrete Element makes this coupling strategy an efficient numerical procedure for the simulation of particle-fluid systems~\shortcite{Cook2004}. In order to capture the actual physical behaviour of the fluid-particle system, the boundary condition between the fluid and the particle is modelled as a non-slip boundary condition, i.e. the fluid near the particle should have similar velocity as the particle boundary. The solid particles inside the fluid are represented by lattice nodes. The discrete nature of lattice will result in stepwise representation of the surfaces. Very small lattice spacing is adopted to obtain smoother boundaries.

@@

\citeNP{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), is compared with a loose
granular column (79\% packing fraction). The permeability is varied by changing
the hydrodynamic radius from 0.7 \textit{R} to 0.95 \textit{R}.

The normalised run-out for different hydrodynamic radii for a granular column 
with an initial aspect ratio of 0.8 for dense initial packing are presented 
in~\cref{fig:Runout_a08_dense}. The run-out increases with decrease in the 
permeability, which is equivalent to an increase in the hydrodynamic radius. 
An increase in the hydrodynamic radius from 0.7 to 0.95 \textit{R} increases 
the normalised run-out by 25\%. However, even under a very low permeability 
condition (\textit{r} = 0.95 \textit{R}), the run-out observed in fluid is 
shorter than the dry and the buoyant conditions.

@@

At a high permeability (\textit{r} = 0.7 \textit{R}), the evolution of run-out 
at the initial stage is quicker, which means that the negative pore-pressure 
that is developed during the shearing along the shear-failure surface is 
dissipated faster. Even though the negative pore-pressure is dissipated, due to 
the development of negative pore-pressure the evolution of run-out in fluid is 
slower than its dry counterpart. The rate of pore-pressure dissipation 
decreases with decrease in the permeability. This can be observed by a flatter 
slope in the run-out evolution with decrease in the 
permeability.~\Cref{fig:PWP_ini_dense} shows the distribution of pore-pressure 
in high and low permeability granular media along the horizontal direction at a 
height of $10 \times d$ from the base. In LBM, the pore-pressure in fluid is a 
function of fluid density distribution functions. At time $t = \tau_c$, the 
highly permeable (\textit{r} = 0.7 \textit{R}) granular column 
shows smaller negative pore-pressure in comparison to large negative 
pore-pressures observed in the shearing zone of a low permeable column 
(\textit{r} = 0.9 R). This shows that not only does it take longer for the 
pore-pressure to dissipate with a decrease in permeability, but also results in 
almost twice the negative pore-pressure than what is observed in the high 
permeable case (\cref{fig:r095_PWP_ini_dense,fig:r095_PWPress_ini_dense}). A 
high value of positive pore-pressure is observed at the face of the column 
in the low permeable condition. This indicates that the low permeable column 
fails as a continuous block undergoing a shear failure, which generates a very 
large negative pore-pressure along the failure surface. However, in the case of 
the high permeable column the failure is more localised with multiple negative 
pore-pressure spikes (\cref{fig:r07_PWPress_ini_dense}).

\begin{figure}
\centering
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r07_PWP_ini_dense}
\end{subfigure} \\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r095_PWP_ini_dense}
\end{subfigure}
\end{figure}
\begin{figure}
\ContinuedFloat
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_press_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r07_PWPress_ini_dense}
\end{subfigure} \\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_press_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r095_PWPress_ini_dense}
\end{subfigure}

\caption{Effect of permeability on the excess pore water pressure distribution 
for a granular column collapse in fluid (a = 0.8 \& dense packing) at $t = 
\tau_c$ along the horizontal direction at a height of 10d from the base.}
\label{fig:PWP_ini_dense}
\end{figure}

Although the low permeability granular columns require a longer duration for 
the run-out to evolve, the final run-out distance is found to be much longer 
than in the high permeable condition.~\Cref{fig:PE_a08_dense} shows that the 
potential energy mobilised during the flow of a low permeability column is 20\% 
smaller than the collapse of a high permeability granular column. Normalised 
kinetic energy evolution with time (\cref{fig:a08_dense_energy}) shows that the 
low permeability column has a wider peak kinetic energy distribution in 
comparison to a sharp peak observed in the high permeability condition. This 
indicates the influence of lubrication, i.e., hydroplaning of the granular flow 
in low permeability conditions. The evolution of the horizontal kinetic energy 
with time reveals that the peak kinetic energy is sustained longer as the 
permeability of the granular material decreases (\cref{fig:KExy_a08_dense}). 
Although the peak kinetic energy is smaller in the low permeability case, the 
hydroplaning of the flowing granular mass results in a longer run-out 
distance.~\Cref{fig:r095_PWP_flow_dense,fig:r095_PWPress_flow_dense} shows the 
distribution of pore-pressure for a dense granular column collapse in fluid 
along the bottom plane. A high positive pore-pressure is observed 
at the base of the granular flow at the flow front in low permeability 
condition indicating the occurrence of hydroplaning. The positive pore-pressure 
at the flow front decreases the effective stress thus the creating lubrication 
effect. The evolution of local packing density with time shows that the packing 
density decreases with decrease in permeability 
(\cref{fig:Packing_Density_a08_dense}). This drop 
in the value of packing density between $t = 2\tau_c$ and $t=3\tau_c$ 
corroborates with the duration of hydroplaning during which a large amount of 
water in entrained at the flow front.

\begin{figure}
	\centering
    \includegraphics[width=0.7\linewidth]{figs/PE_a08_dense}
    \caption{Effect of permeability on the evolution of the potential energy 
    with time for a granular column collapse in fluid (a = 0.8).}
    \label{fig:PE_a08_dense}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.8\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KE_a08_dense}
    \caption{Evolution of the total kinetic energy.}
    \label{fig:KE_a08_dense}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KExy_a08_dense}
    \caption{Evolution of horizontal and vertical kinetic energies.}
    \label{fig:KExy_a08_dense}
\end{subfigure}
}
\caption{Effect of permeability on the evolution of kinetic energies with time 
for a granular column collapse in fluid (a = 0.8).}
\label{fig:a08_dense_energy}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
    bottom of the granular flow.}
    \label{fig:r07_PWP_flow_dense}
\end{subfigure}
\\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r095_PWP_flow_dense}
\end{subfigure}
\end{figure}
\begin{figure}
\ContinuedFloat
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_press_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r07_PWPress_flow_dense}
\end{subfigure}
\\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_press_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r095_PWPress_flow_dense}
\end{subfigure}
\caption{Effect of permeability on the excess pore water pressure distribution 
along the bottom plane for a granular column collapse in fluid (a = 0.8 \& 
dense packing) at $t = 2\tau_c$.}
\label{fig:PWP_flow_dense}
\end{figure}

@@

The high permeability granular column shows lower water entrainment 
(\cref{fig:Packing_Density_a08_dense}), which indicates that for highly 
permeable flows the drag force acting on the soil grains predominates over the 
lubrication effect on the run-out behaviour.~\Cref{fig:Fr_a08_dense} shows the 
evolution of Froude's number with time for different permeability conditions. 
In dense granular columns, the Froude's number increases with decrease in 
permeability, however the Froude's number is below a critical value of 0.4 even 
for a low permeable column. Hence, no hydroplaning is observed. In both the low 
and high permeable granular flows, the granular material consolidates at the 
final stage of the flow (\cref{fig:Packing_Density_a08_dense}). This can be 
observed by the increase in the packing density at the final stage due to 
settlement of grains and expulsion of entrained water. The final deposit 
profile for both low and high permeability conditions are shown 
in~\cref{fig:a08_dense_snapshots}. The high permeability collapse shows a more 
parabolic (convex) deposit profile in contrast to the more concave profile 
observed in the low permeability condition. In high permeable condition, once 
the granular mass is mobilised, the drag forces slows down the rate of flow, 
which allows the grains to densify resulting in a more parabolic profile. This 
is in contrast to the hydroplaning observed in the low permeability condition, 
where the granular mass thickness is smaller than the dense condition. The 
observation of hydroplaning in the low permeable condition may be due to the 
difference in the distribution of granular mass at the flow front.

@@

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r07_PWP_ini}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r07_PWP_ini}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r095_PWP_ini}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
    bottom of the granular flow.}
    \label{fig:Loose_r095_PWP_ini}
\end{subfigure}
}
\end{figure}
\begin{figure}
\ContinuedFloat
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_press_loose}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r07_PWPress_ini}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_press_loose}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r095_PWPress_ini}
\end{subfigure}
}
\caption{Effect of permeability on the excess pore water pressure distribution 
along the base of a granular column collapse in fluid (a = 0.8 \& loose 
packing) at $t = \tau_c$.}
\label{fig:Loose_PWP_ini}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r07_PWP_flow}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r07_PWP_flow}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r095_PWP_flow}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r095_PWP_flow}
\end{subfigure}
}
\end{figure}
\begin{figure}
\ContinuedFloat
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_press_loose}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r07_PWPress_flow}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_press_loose}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r095_PWPress_flow}
\end{subfigure}
}
\caption{Effect of permeability on the excess pore water pressure distribution 
for a granular column collapse in fluid (a = 0.8 \& loose packing) at $t = 
2\tau_c$.}
\label{fig:Loose_PWP_flow}
\end{figure}

The evolution of the potential energy with time (\cref{fig:PE_a08_loose}) 
reveals that at a very low permeability (\textit{r} = 0.95 \textit{R}), the 
initial potential energy mobilised is smaller than at \textit{r} = 0.9 
\textit{R}. Also with decreasing permeability, the time required to 
dissipate the negative pore-pressure increases. This results in a shorter 
run-out distance in the case of \textit{r} = 0.95 \textit{R} to that of 
\textit{r} = 0.9 \textit{R}. As the quantity of material destabilised is small, 
the flow is thinner and thus has a high Froude's number 
(0.59).~\Cref{fig:KExy_a08_loose} shows that the peak horizontal kinetic 
velocity observed in the case of \textit{r} = 0.9 \textit{R} is higher than 
\textit{r} = 0.95 \textit{R}. A Froude's number of 0.59 for \textit{r} = 0.9 
\textit{R} is observed in contrast to 0.46 for \textit{r} = 0.95 \textit{R}. 
Both values of hydrodynamic radii result in a Froude's number that indicate the 
occurrence of hydroplaning. However, the difference in the amount of material 
destabilised for \textit{r} = 0.95 \textit{R} and the decreased effect of 
hydroplaning results in a shorter run-out distance for \textit{r} = 0.95 
\textit{R} in comparison to \textit{r} = 0.9 \textit{R}.

@@

\begin{figure}
	\centering
    \includegraphics[width=0.8\linewidth]{figs/PE_a08_loose}
    \caption{Effect of permeability on the evolution of the potential energy 
    with time for a granular column collapse in fluid (a = 0.8 \& loose 
    packing).}
    \label{fig:PE_a08_loose}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.8\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KE_a08_loose}
    \caption{Evolution of the total kinetic energy.}
    \label{fig:KE_a08_loose}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KExy_a08_loose}
    \caption{Evolution of horizontal and vertical kinetic energies.}
    \label{fig:KExy_a08_loose}
\end{subfigure}
}
\caption{Effect of permeability on the evolution of kinetic energies with time 
for a granular column collapse in fluid (a = 0.8 \& loose packing).}
\label{fig:a08_loose_energy}
\end{figure}

@@

The evolution of grain trajectories with time are presented 
in~\cref{fig:Loose_a08_permeability} for low (\textit{r} = 0.95 \textit{R}) and 
high (\textit{r} = 0.9 \textit{R})
permeability conditions. It can be observed that a high 
permeability column shows a parabolic (convex) final profile in contrast to the 
more concave profile observed in low permeability condition, due to the effect 
of drag forces on the flow front. This difference in the flow thickness results 
in a higher value of Froude's number (0.59) and the occurrence of hydroplaning 
in the low permeability condition. Due to the high permeability, the water 
entrained at the flow front is dissipated quicker and 
thus no lubrication effect is observed. A Froude's number of 0.272 (no 
hydroplaning) is observed for the high permeability condition (\textit{r} = 0.7 
\textit{R}). The 
thick flow front in dense condition results in higher effective stress, in 
contrast to the low effective stress in loose condition due to positive 
pore-pressure at the flow front. The higher effective stress results in more 
frictional dissipation in dense condition, while the loose column experiences
lubrication effect. This shows that the drag force predominates at high 
permeability, while the low permeability condition is characterised by 
hydroplaning and lubrication.

\Cref{fig:effective_stress_a08} shows the normalised pressure at the 
base for the low and high permeability flows at $ t = 2\tau_c$. The normalised 
effective stress plotted is obtained as the average over 5 time steps at 
$2\tau_c$. The effective stress at the base is normalised to the effective 
stress of a static granular column before the collapse. A value of 1 indicates 
that the effective stress hasn't changed, which can be observed in the static 
region of the granular column. It can be observed that the normalised effective 
stress is significantly higher for the high permeability condition at the flow 
front in comparison to the almost non-existence of effective stress in the low 
permeability condition. The observation of trivial effective stress at the flow 
front corroborates the lubrication effect observed at low permeability 
conditions.

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/LBM_MD_loose_a08_r07_pt}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}).}
    \label{fig:LBM_MD_loose_a08_r07_pt}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/LBM_MD_a08_Loose_r09}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}).}
    \label{fig:LBM_MD_a08_Loose_r09}
\end{subfigure}
}
\caption{Particle tracking of the deposit morphology
for a granular column collapse in fluid (a = 0.8 \& loose packing), influence 
of permeability.}
\label{fig:Loose_a08_permeability}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.97\columnwidth]{figs/a08/effective_stress_a08}
\caption{Effect of permeability on the normalised effective stress for loose 
initial packing at $t = 2\tau_c$.}
\label{fig:effective_stress_a08}
\end{figure}

>>>>

\section{COUPLED LB - DEM FOR FLUID-PARTICLE INTERACTIONS}
The Lattice Boltzmann approach has the advantage of accommodating large particle sizes and the interaction between the fluid and the moving particles can be modelled through relatively simple fluid - particle interface treatments. Further, employing the Discrete Element Method (DE) to account for the particle/particle interaction naturally leads to a combined LB - DEM solution procedure. The Eulerian nature of the Lattice Boltzmann formulation, together with the common explicit time step scheme of both the Lattice Boltzmann and the Discrete Element makes this coupling strategy an efficient numerical procedure for the simulation of particle-fluid systems~\shortcite{Cook2004}. In order to capture the actual physical behaviour of the fluid-particle system, the boundary condition between the fluid and the particle is modelled as a non-slip boundary condition, i.e. the fluid near the particle should have similar velocity as the particle boundary. The solid particles inside the fluid are represented by lattice nodes. The discrete nature of lattice will result in stepwise representation of the surfaces. Very small lattice spacing is adopted to obtain smoother boundaries. Further details about the modelling can be found in~\shortciteNP{Kumar2015,Kumar2017}.

@@

\shortciteNP{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), is compared with a loose
granular column (79\% packing fraction). The permeability is varied by changing
the hydrodynamic radius from 0.7 \textit{R} to 0.95 \textit{R}.

\subsection{Dense granular collapse}
The normalised run-out for different hydrodynamic radii for a granular column 
with an initial aspect ratio of 0.8 for dense initial packing are presented 
in~\cref{fig:Runout_a08_dense}. The run-out increases with decrease in the 
permeability, which is equivalent to an increase in the hydrodynamic radius. 
An increase in the hydrodynamic radius from 0.7 to 0.95 \textit{R} increases 
the normalised run-out by 25\%. However, even under a very low permeability 
condition (\textit{r} = 0.95 \textit{R}), the run-out observed in fluid is 
shorter than the dry and the buoyant conditions.

@@

At a high permeability (\textit{r} = 0.7 \textit{R}), the evolution of run-out 
at the initial stage is quicker, which means that the negative pore-pressure 
that is developed during the shearing along the shear-failure surface is 
dissipated faster. Even though the negative pore-pressure is dissipated, due to 
the development of negative pore-pressure the evolution of run-out in fluid is 
slower than its dry counterpart. The rate of pore-pressure dissipation 
decreases with decrease in the permeability. This can be observed by a flatter 
slope in the run-out evolution with decrease in the 
permeability.~\Cref{fig:PWP_ini_dense} shows the distribution of pore-pressure 
in high and low permeability granular media along the horizontal direction at a 
height of $10 \times d$ from the base. In LBM, the pore-pressure in fluid is a 
function of fluid density distribution functions. At time $t = \tau_c$, the 
highly permeable (\textit{r} = 0.7 \textit{R}) granular column 
shows smaller negative pore-pressure in comparison to large negative 
pore-pressures observed in the shearing zone of a low permeable column 
(\textit{r} = 0.9 R). This shows that not only does it take longer for the 
pore-pressure to dissipate with a decrease in permeability, but also results in 
almost twice the negative pore-pressure than what is observed in the high 
permeable case (\cref{fig:r095_PWP_ini_dense}). A 
high value of positive pore-pressure is observed at the face of the column 
in the low permeable condition. This indicates that the low permeable column 
fails as a continuous block undergoing a shear failure, which generates a very 
large negative pore-pressure along the failure surface. However, in the case of 
the high permeable column the failure is more localised with multiple negative 
pore-pressure spikes (\cref{fig:r07_PWP_ini_dense}).

\begin{figure}
\centering
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r07_PWP_ini_dense}
\end{subfigure} \\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r095_PWP_ini_dense}
\end{subfigure}
\caption{Effect of permeability on the excess pore water pressure distribution 
for a granular column collapse in fluid (a = 0.8 \& dense packing) at $t = 
\tau_c$ along the horizontal direction at a height of 10d from the base.}
\label{fig:PWP_ini_dense}
\end{figure}

Normalised kinetic energy evolution with time (\cref{fig:KExy_a08_dense}) shows that the 
low permeability column has a wider peak kinetic energy distribution in 
comparison to a sharp peak observed in the high permeability condition. This 
indicates the influence of lubrication, i.e., hydroplaning of the granular flow 
in low permeability conditions. The evolution of the horizontal kinetic energy 
with time reveals that the peak kinetic energy is sustained longer as the 
permeability of the granular material decreases . 
Although the peak kinetic energy is smaller in the low permeability case, the 
hydroplaning of the flowing granular mass results in a longer run-out 
distance.~\Cref{fig:r095_PWP_flow_dense} shows the 
distribution of pore-pressure for a dense granular column collapse in fluid 
along the bottom plane. A high positive pore-pressure is observed 
at the base of the granular flow at the flow front in low permeability 
condition indicating the occurrence of hydroplaning. The positive pore-pressure 
at the flow front decreases the effective stress thus the creating lubrication 
effect. The evolution of local packing density with time shows that the packing 
density decreases with decrease in permeability 
(\cref{fig:Packing_Density_a08_dense}). This drop 
in the value of packing density between $t = 2\tau_c$ and $t=3\tau_c$ 
corroborates with the duration of hydroplaning during which a large amount of 
water in entrained at the flow front.

\begin{figure}
    \includegraphics[width=\linewidth]{figs/KExy_a08_dense}
    \caption{Effect of permeability on the evolution of kinetic energies with time 
for a granular column collapse in fluid (a = 0.8).}
    \label{fig:KExy_a08_dense}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
    bottom of the granular flow.}
    \label{fig:r07_PWP_flow_dense}
\end{subfigure}
\\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r095_PWP_flow_dense}
\end{subfigure}
\caption{Effect of permeability on the excess pore water pressure distribution 
along the bottom plane for a granular column collapse in fluid (a = 0.8 \& 
dense packing) at $t = 2\tau_c$.}
\label{fig:PWP_flow_dense}
\end{figure}

@@

The high permeability granular column shows lower water entrainment 
(\cref{fig:Packing_Density_a08_dense}), which indicates that for highly 
permeable flows the drag force acting on the soil grains predominates over the 
lubrication effect on the run-out behaviour.~\Cref{fig:Fr_a08_dense} shows the 
evolution of Froude's number with time for different permeability conditions. 
In dense granular columns, the Froude's number increases with decrease in 
permeability, however the Froude's number is below a critical value of 0.4 even 
for a low permeable column. Hence, no hydroplaning is observed. In both the low 
and high permeable granular flows, the granular material consolidates at the 
final stage of the flow (\cref{fig:Packing_Density_a08_dense}). This can be 
observed by the increase in the packing density at the final stage due to 
settlement of grains and expulsion of entrained water.

\subsection{Loose granular collapse}

@@

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r07_PWP_ini}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r07_PWP_ini}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r095_PWP_ini}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
    bottom of the granular flow.}
    \label{fig:Loose_r095_PWP_ini}
\end{subfigure}
}
\caption{Effect of permeability on the excess pore water pressure distribution 
along the base of a granular column collapse in fluid (a = 0.8 \& loose 
packing) at $t = \tau_c$.}
\label{fig:Loose_PWP_ini}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r07_PWP_flow}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r07_PWP_flow}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r095_PWP_flow}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r095_PWP_flow}
\end{subfigure}
}
\caption{Effect of permeability on the excess pore water pressure distribution 
for a granular column collapse in fluid (a = 0.8 \& loose packing) at $t = 
2\tau_c$.}
\label{fig:Loose_PWP_flow}
\end{figure}

The initial potential energy mobilised is smaller than at \textit{r} = 0.9 
\textit{R}. Also with decreasing permeability, the time required to 
dissipate the negative pore-pressure increases. This results in a shorter 
run-out distance in the case of \textit{r} = 0.95 \textit{R} to that of 
\textit{r} = 0.9 \textit{R}. As the quantity of material destabilised is small, 
the flow is thinner and thus has a high Froude's number 
(0.59).~\Cref{fig:KExy_a08_loose} shows that the peak horizontal kinetic 
velocity observed in the case of \textit{r} = 0.9 \textit{R} is higher than 
\textit{r} = 0.95 \textit{R}. A Froude's number of 0.59 for \textit{r} = 0.9 
\textit{R} is observed in contrast to 0.46 for \textit{r} = 0.95 \textit{R}. 
Both values of hydrodynamic radii result in a Froude's number that indicate the 
occurrence of hydroplaning. However, the difference in the amount of material 
destabilised for \textit{r} = 0.95 \textit{R} and the decreased effect of 
hydroplaning results in a shorter run-out distance for \textit{r} = 0.95 
\textit{R} in comparison to \textit{r} = 0.9 \textit{R}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figs/KExy_a08_loose}
    \caption{Effect of permeability on the evolution of kinetic energies with time 
for a granular column collapse in fluid (a = 0.8 \& loose packing).}
    \label{fig:KExy_a08_loose}
\end{figure}

@@
